---
title: "W2_L05_Correlation Coding Lecture"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Correlation.

# How to quantify similarity between variables?

## Read data and explore it

```{r}
library(tidyverse, quietly=TRUE)
ds1 <- read.csv("data/DATA_FSB_SET_1.csv", row.names = 1)
glimpse(ds1)
```

## Visual exploration

A scatterplot suggests a relationship between our variables Hours_Sun and LDL.

```{r}
ggplot(ds1, aes(x=Hours_Sun, y=LDL)) + 
    geom_point()
```

## How do we quantify such a relationship? Covariance

´Manual´ calculation of covariance

```{r}
Hours_sun_LDL_cov <- sum((ds1$Hours_Sun - mean(ds1$Hours_Sun)) * (ds1$LDL - mean(ds1$LDL))) / (nrow(ds1) -1)
round(Hours_sun_LDL_cov, 2)
```

## How do we quantify such a relationship? Covariance

R has a built-in function for the covariance: cov

```{r}
Hours_sun_LDL_cov <- cov(ds1$Hours_Sun, ds1$LDL)
round(Hours_sun_LDL_cov, 2)

```

# Computing the correlation

## Manual calculation of correlation

```{r}
Hours_sun_LDL_cov <- sum((ds1$Hours_Sun - mean(ds1$Hours_Sun)) * (ds1$LDL - mean(ds1$LDL))) / (nrow(ds1) -1)
Hours_sun_LDL_cor <- Hours_sun_LDL_cov / (sd(ds1$Hours_Sun) * sd(ds1$LDL))
cat("Pearson correlation Hours_sun vs. LDL: ", round(Hours_sun_LDL_cor, 3), '\n')
```

## R built-in function for the correlation

```{r}
Hours_sun_LDL_cor <- cor(ds1$Hours_Sun, ds1$LDL)
```

## Statistical analysis.

The alternative hypothesis is defined as the correlation being different to 0

```{r}
cor.test(ds1$Hours_Sun, ds1$LDL)
```

## Statistical analysis: alternative tests.

We can specify the alternative hypothesis, e.g. that the correlation is larger than 0.

```{r}
cor.test(ds1$Hours_Sun, ds1$LDL, alternative = 'greater')
```

# Assumptions for correlation.

## Gaussian/Normal distributions for each variable.

```{r}
shapiro.test(ds1$Hours_Sun)
shapiro.test(ds1$LDL)
```

## Non-parametric correlation analysis.

```{r}
cor.test(ds1$Hours_Sun, ds1$LDL,method="spearman")
```
