---
title: "W2_L06_Linear models Coding Lecture"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Linear models.

# Some overview

Let's fit a model to our observed relationship between Weight and the
LDL concentration and inspect the result.

## General format:

lm(, )

lm(Response \~ Predictor, data)

## Assumptions to allow fitting a linear model:

-   Dependency between data is linear
-   Mean of residuals is zero
-   Homoscedasticity of residuals or equal variance
-   No autocorrelation of residuals
-   The X variables and residuals are uncorrelated
-   The number of observations must be greater than number of Xs
-   The variability in X values is positive
-   The regression model is correctly specified
-   No perfect multicollinearity
-   Normality of residuals

# Linear model

```{r}
library(tidyverse)
```

## First load our trusty dataset 1

```{r}
ds1 <- read.csv("data/DATA_FSB_SET_1.csv", row.names = 1)
```

## Fit a linear model

```{r}
wt.vs.ldl.lm <- lm(Weight ~ LDL, ds1)
```

## Print the summary of the model

```{r}
summary(wt.vs.ldl.lm)
```

## What we should be looking into...

-   The estimated intercept and slope are -156.8 and 1.9.

-   The result looks good, the adjusted R-square value of 0.9992 is
    close to 1.

-   The definition of R-squared is the percentage of the response
    variable variation that is explained by a linear model.

    ```         
    R-squared = Explained variation / Total variation 
    R-squared is always between 0 and 1.
    ```

## Access the values

```{r}
s <- summary(wt.vs.ldl.lm)
s$r.squared
```

# Evaluate the linear model

## Let's inspect the result visually.

First with a simple plot of our two variables, and the linear fit.

```{r}
plt <- ds1 %>% 
  ggplot(aes(x = LDL, y = Weight)) + 
  geom_point() +
  xlab("LDL") +
  ylab("Weight") +
geom_smooth(method='lm', se = FALSE)  
plt
```

As not all data points are perfectly captured by the linear model, we
should inspect the residuals with respect to the fitted values.

## Distribution of the residuals

```{r}
mean(wt.vs.ldl.lm$residuals)

hist(wt.vs.ldl.lm$residuals)
```

## QQ-plot

Comparing the residual distribution to a theoretical normal distribution
in a quantile-quantile-plot (QQ-plot).

```{r}
plot(wt.vs.ldl.lm, 2)
```

## Fitted value vs Residuals

Another intuitive way is to examine the relationship between the fitted
value and the corresponding residual. We see that the model
systematically deviates from the data, overestimating the weight for
very low and high LDL values, and slightly underestimates it for
intermediate values.

```{r}
options(repr.plot.width = 6, repr.plot.height = 6)
plot(wt.vs.ldl.lm, 1)
```
