---
title: "W2_L02_Logical Vectors Coding Lecture"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
# Note that the `echo = FALSE` parameter was added to the code chunk 
# to prevent printing of the R code that generated the plot.
```

# WEEK 2 LECTURE 2.

# Operations with logical vectors

## Logical vectors contain only boolean values

{TRUE, FALSE}

## Boolean values can be abbreviated

TRUE = T FALSE = F

## Boolean values are the result of a range of logical operators:

== equal != not equal \< less than \<= less than or equal \> greater
than \>= greater than or equal & and \| or ! not %in% contained in set

## Examples

```{r}
1 == 2
1 < 2
!TRUE
1 %in% c(1,2,3)
```

## Examples with vectors

```{r}
v <- 1:10
v > 5
v >= 6
```

# Indexing a data structure

## Logical vectors can be used to select certain parts from data structures like data frames or matrices.

```{r}
mat <- matrix(data = rnorm(40), nrow = 4, ncol = 10)
dim(mat)

idx <- c(TRUE, TRUE, TRUE, FALSE)

mat2 <- mat[idx,]
dim(mat2)
```

## Working with a real data-set

```{r}
# Set working directory 
setwd(".")

# Load dataset
ds1 <- read.csv("data/DATA_FSB_SET_1.csv", row.names = 1)

# Check its dimensions. Check out the documentation with "?dim"
dim(ds1)
```

```{r}
# Lets have a look to the data
head(ds1)
```

```{r}
# Access third, fifth, and 99th row directly.
ds1[c(3,5,99), ]
```

## Aim: identify samples of interest.

### But we can use our logical operators to make a selection index.

### question 1: which patients are more than 100 kg in weight?

### question 2: store that information into a vector.

### question 3: explore the results.

```{r}

# identify and store
ds1$Weight > 100
idx <- ds1$Weight > 100

# explore the results.
table(idx)
sum(idx)
```

## Aim: select samples of interest.

### remember the code

```{r}
rownames(ds1)
ds1$Weight > 100
```

### Lets use the vector to filter the rownames of interest.

```{r}
rownames(ds1)[ ds1$Weight > 100 ]
```

## Aim: how many patients are experiencing in average exactly 10 hours of sun per day?

```{r}
table(ds1$Hours_Sun == 10)
```

## Aim: compare two columns

```{r}
table(ds1$Hours_Sun > ds1$Sleep_Hours)
```

## Or how many patients have been in the hospital from 2 to 4 times?

```{r}
table(ds1$Hospital_Visits %in% 2:4)
```

# Combinations of logical operations

But we can also ask more complex questions, using boolean algebra. This
allows chaining multiple logical operations. This is done by the
operators "and" (&) and "or" (\|), which take two boolean values.

TRUE & TRUE TRUE \| FALSE

## Map of combinations

The results of all possible combinations of boolean value and operator
is shown in this truth table:

| X \| Y \| "X & Y" \| "X \| Y" |
|-------------------------------|
| T \| F \| F \| T              |
| T \| F \| F \| T              |
| T \| T \| T \| T              |
| F \| F \| F \| F              |

## Making use of the combinations.

Armed with that logic, we can connect conditions, like rewriting the
previous range filter.

```{r}
table(ds1$Hospital_Visits >= 2 & ds1$Hospital_Visits <= 4)
```

Or connect conditions on multiple columns, to e.g. find genes that have
large count numbers in both samples.

```{r}
table(ds1$Hospital_Visits >= 4 & ds1$Weight > 100)
```

These conditions can be arbitrarily complex, such as finding patients
which are more than 100 kg in weight with more than 3 hospitalizations
or are older than 80 years of age.

```{r}
table((ds1$Hospital_Visits >= 4 & ds1$Weight > 100) | ds1$Age >= 80)
```

# Application of logical vectors to filter data

A simple but very powerful of logical vectors is filtering tables. Lets
follow an example where we will filter unrealistic values.

##Lets start with a plot. Which we will see how to do more of those in
the coming lectures. A histogram allows a simple yet powerful data
exploration

```{r}
hist(ds1$Hours_Sun,breaks=10)
```

After we observed the plot, we identified several individuals with very
few hours of sun. Those values are to be investigated. Maybe they are
living in countries with very few hours of sun, or they work mainly
indoors or there are other possible reasons. Let us consider for this
exercise that those are wrongly measured values, and - as a result - we
decide to filter the values

## First we need to identify those individuals. Let us consider that less than 3 hours is not a realistic value.

```{r}
ds1$Hours_Sun < 3
ind_wrong_values <- ds1$Hours_Sun < 3
```

## In a second step we, we are making use of the logical vector to filter

```{r}
ds1_filt <- ds1[!ind_wrong_values,]
dim(ds1_filt)
```

We could decide to extend this filtering to additional variables. For
instance, to filter outliers.
