---
title: "W2_L07_t-test_ANOVA Coding Lecture"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# t-test

## Aim

Statistical test to compare the means of two populations.

## Assumptions

-   The values from each group follow a normal distribution.
-   Homogeneous, or equal, variance exists when the standard deviations
    of samples are approximately equal.
-   Scale of measurement applied to the data collected follows a
    continuous or ordinal scale.
-   Data is collected from a representative, randomly selected portion
    of the total population.

## Load data

```{r}
# Use dataset 3
# Consider that patients from earth have a different LDL level than patients from venus
ds3 <- read.csv("data/DATA_FSB_SET_3.csv", row.names = 1)
summary(ds3)
# NO NAs...
```

```{r}
# install.packages("tidyverse")
# install.packages("corrr")
library(tidyverse)
library(corrr)

ds3_corr <- ds3 %>% 
  select(-Planet) %>% 
  select(-Planet2) %>% 
  cor(method = "pearson")

# plot sample correlation structure as heatmap
rplot(ds3_corr) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Visual inspection with box plots shows some difference between the planets
ds3.df <- ds3 %>%
mutate(Planet = factor(Planet)) %>%
mutate(Planet2 = factor(Planet2))

# Let's have a look at the resulting dataframe
glimpse(ds3.df)
```

```{r}
ds3.df %>%
group_by(Planet) %>%
summarise(
count_planet = n(),
mean_LDL = mean(LDL_levels, na.rm = TRUE),
sd_LDL = sd(LDL_levels, na.rm = TRUE)
)

plt <- ggplot(ds3.df, aes(x = Planet, y = LDL_levels, fill = Planet)) +
geom_boxplot() +
theme_classic()
plt
```

## Analyze assumptions

Before we can employ a t-Test, we have to assess whether the fundamental
assumptions hold.

### Asumption 1: each group follows a normal distribution.

A histogram provides a useful visualization. Let's look at LDL level
histogram:

```{r}
# focus on Venus, since we can hypothesize that there is a difference between the planets
planet <- 'Venus'

# ggplot2 allows to easily fit a normal distribution on our data for a visual assessment
plt <- ds3 %>% 
filter(Planet == planet) %>% 
    ggplot() + 
    geom_histogram(aes(x = LDL_levels, y = ..density.., fill = Planet), binwidth = 5) +
    xlab("LDL [mg/dL]") +
    stat_function(fun = dnorm, 
                  args = list(mean = mean(ds3$LDL_levels[ds3$Planet == planet]), 
                              sd = sd(ds3$LDL_levels[ds3$Planet == planet]))) +
scale_fill_manual(values=c("#E69F00", "#56B4E9","#999999"))
plt
```

But visual assessment can be deceiving. A more precise way is a
quantile-quantile plot:

```{r}
options(repr.plot.width = 7, repr.plot.height = 4)
par(mfrow = c(1,2), bg = 'white')
# For Earth
qqnorm(ds3$LDL_levels[ds3$Planet == 'Earth'], pch = 1, frame = FALSE, main = 'Earth')
qqline(ds3$LDL_levels[ds3$Planet == 'Earth'], col = "steelblue", lwd = 2)

# For Venus
qqnorm(ds3$LDL_levels[ds3$Planet == 'Venus'], pch = 1, frame = FALSE, main = 'Venus')
qqline(ds3$LDL_levels[ds3$Planet == 'Venus'], col = "steelblue", lwd = 2)
```

```{r}
shapiro.test(ds3$LDL_levels[ds3$Planet == 'Earth'])
shapiro.test(ds3$LDL_levels[ds3$Planet == 'Venus'])
```

### Asumption 2: each group has the same variance.

The Bartlett test can be used to determine if the variance is similar
between two samples.

```{r}
ldl_vs_planet.varTest <- bartlett.test(LDL_levels ~ Planet, data = ds3)
ldl_vs_planet.varTest
```

### Asumption 3: Independence

The third assumption is independence of samples from each other. This
assumption needs to be investigated in the context of the experimental
design. Let us assume, individuals are coming from different families,
and there is no bias towards selection (e.g. not a bias towards
younger).

## Running the t-test

```{r}
# first without variance-equality option
ldl.vs.planet.ttest <-  t.test(LDL_levels ~ Planet, data=ds3)
ldl.vs.planet.ttest
# the actual p-value
signif(ldl.vs.planet.ttest$p.value, digits=3)
```

```{r}
# second with variance-equality option, since we tested that
ldl.vs.planet.ttest.2 <-  t.test(LDL_levels ~ Planet, data=ds3, var.equal = TRUE)
ldl.vs.planet.ttest.2
# the actual p-value
signif(ldl.vs.planet.ttest.2$p.value, digits=3)
```

As a result, considering that we set a threshold for significance at
0.01, we can reject the null hypothesis, and we will consider the two
populations to have a different mean.

# ANOVA

Let us review the comparisons between means when instead of two groups,
we have three groups.

## Comparing the means.

```{r}
ds3 %>%
group_by(Planet2) %>%
summarise(
count_planets = n(),
mean_time = mean(LDL_levels, na.rm = TRUE),
sd_time = sd(LDL_levels, na.rm = TRUE)
)
```

## Comparing the empirical distributions between the planets using boxplots

```{r}
#Lets look at the 
plt <- ggplot(ds3.df, aes(x = Planet2, y = LDL_levels, fill = Planet2)) +
geom_boxplot() +
theme_classic() +
theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
plt
```

## Running the ANOVA test

```{r}
# Now for the systematic analysis using a one way ANOVA
anova_one_way <- aov(Exercise ~ Planet2, data = ds3)
summary(anova_one_way)
```

## Post-hoc analysis

```{r}
TukeyHSD(anova_one_way, ordered = TRUE)
plot(TukeyHSD(anova_one_way, ordered = TRUE))
```
