---
title: "Data Analysis: case study"
author: "Robert Lehmann (modified by David Gomez-Cabrero)"
date: '2022'
output: html_document
---

# 0- Having a plan.

a.  Load the data.
b.  Check NA, etc...
c.  Explore the data: scatter-plots, box-plots, etc...
d.  Investigate the possible association(s) using statistics
e.  Initial conclusions

# 1- Setting the environment

```{r}

getwd()

#setwd()

## YOU NEED TO SET UP YOUR WORKING DIRECTORY

dir()
```

# 2- Load the data and initial exploration

```{r}
library(tidyverse)
# What happens if this package is not installed in my library?
# install.packages("tidyverse")
library(corrr)

# Load dataset 4 A
ds4 <- read.csv("r-data/DATA_FSB_SET_4A.csv", row.names = 1)
# quick check if there are any missing values
any(!complete.cases(ds4))
ds4 <- as.matrix(ds4)
head(ds4)
```

# 3- Investigating the distribution of the data.

```{r}
# Looking at the total data distribution reveals a non-normal distribution.
hist(ds4, xlab = 'nominal expression values')
# Expression data are commonly log-transformed to achieve a normal distribtution, lets try that
ds4.lg <- log1p(ds4)
hist(ds4.lg, xlab = 'log scale expression values with pseudocount 1')

# lets distinguish between planets
data <- ds4.lg %>% 
    as.data.frame %>%
    pivot_longer(names_to = "patient", values_to = "expression", cols = 1:ncol(ds4))
data$planet <- 'Venus'
data$planet[grep('Earth', data$patient)] <- 'Earth'


plt <- data %>% 
    ggplot() + 
    geom_histogram(aes(x = expression, y = ..density.., fill = planet), 
                   binwidth = .5, alpha=.6) +
    xlab("Expression [log1p]") 
plt
```

# 3- Investigating the distribution of the data.

```{r}
# The combination of all genes does not look like a normal distribution, not too surprising since 
# we expect gene regulation to play a role in causing very different expression patterns between some genes, 
# but also between planets.
# So lets look at the replicated measurements of one gene on one planet. 
# We can again use a quantile-quantile plot.

venus.idx <- grep("Venus",colnames(ds4.lg))
earth.idx <- grep("Earth",colnames(ds4.lg))

qqnorm(ds4.lg[3,venus.idx])
qqline(ds4.lg[3,venus.idx], col = "steelblue", lwd = 2)
```

# 3B- Investigating the distribution of the data.

```{r}
# There are also statistical tests for normality, like the Kolmogorov-Smirnov test or the Shapiro-Wilks test.
# Let's use the latter for this gene, distinguishing between planets. 
shapiro.test(ds4.lg[3,venus.idx])
shapiro.test(ds4.lg[3,earth.idx])
```

# 4- Investigating the distribution of the "samples"

```{r}
ds4.lg.pca <- prcomp(t(ds4.lg), scale = FALSE, center = FALSE)

# the scree plot shows us, which dimensions carry the majority of the variation.
options(repr.plot.width = 6, repr.plot.height = 4)
plot(ds4.lg.pca,
	xlab = "Dimension",
	main = 'Scree plot')

# cumulative explained variability plot
cp <- cumsum(ds4.lg.pca$sdev^2 / sum(ds4.lg.pca$sdev^2))
plot(cp, 
    xlab = "PC #", 
     ylab = "Amount of explained variance", 
     main = "Cumulative variance plot"
    )
# The vast majority of variance in the dataset (98%) is represented in the first principal component!
col.by.planet <- rep('Earth', ncol(ds4.lg)) 
col.by.planet[grep('Venus', colnames(ds4.lg))] <- 'Venus'

# now lets look at whether the patients cluster somehow
library("factoextra")
options(repr.plot.width = 8, repr.plot.height = 8)

fviz_pca_ind(ds4.lg.pca,
             axes = c(1, 2),
             geom = c("point"),
            col.ind = col.by.planet)
```

# 5- Differential analysis: for a single-gene

```{r}
# make index of earth and venus patient samples, respectively
venus.idx <- grep('Venus', colnames(ds4))
earth.idx <- grep('Earth', colnames(ds4))

# now we can e.g. run a t-test to determine if Gene1 expression values are significantly different 
t.test(ds4.lg['Gene1',venus.idx], ds4.lg['Gene1',earth.idx])
```

# 6- Differential analysis: for all genes

```{r}
# It is not practical to do this for each gene, so lets perform a t-test for every gene, but without using a loop
p.vals <- sapply(1:nrow(ds4.lg), 
                 function(i) 
                     t.test(ds4.lg[i, earth.idx],ds4.lg[i, venus.idx])[c("p.value")]
                 )
# how many genes do we find to be significantly different?
table(p.vals < 0.05)
```

# 7- Heatmaps

```{r}
# A visual representation of the presumably differentially expressed genes is very helpful. Let's make a heatmap.
library(pheatmap)

# a logical vector indicating the genes with significant p-values from the t-test
de.idx <- p.vals < 0.05

options(repr.plot.width = 7, repr.plot.height = 20)
pheatmap(ds4.lg[de.idx,])
```

# 7- Investigating the variability in genes.

```{r}
# mean vs variance plot -> show that statistics vary between highly and lowly expressed genes 
plot(apply(ds4.lg[,earth.idx], 1, mean), apply(ds4.lg[,earth.idx], 1, var), 
     xlab = 'Mean expression [log]',
    ylab = 'Expression variance [log]', 
    main = 'Expression Mean vs. Variance for Earth samples',
    pch = 19)


# volcano plot
fc.log <- -log10(apply(ds4.lg[,venus.idx], 1, mean) / apply(ds4.lg[,earth.idx], 1, mean))
col.fc <- rep('black', nrow(ds4.lg))
col.fc[p.vals < 0.01 & fc.log < 0] <- 'red'
col.fc[p.vals < 0.01 & fc.log > 0] <- 'green'

plot(fc.log, -log10(unlist(p.vals)), 
     main = 'Volcano plot',
     xlab = 'mean log expression',
    ylab = 'sd log expression',
    col = col.fc,
    pch = 19)
abline(h = -log10(0.01), v = 0)


```
